---
title: "S3 Appendix"
output: 
  pdf_document:
    number_sections: true
urlcolor: blue
---

This appendix serves a double purpose. On the one hand, it aims to demonstrate
the advantages of using fake/synthetic data to diagnose problems in calibration
algorithms. With synthetic data, we aim to answer: "Is it possible to estimate
the desired parameters with the available data if we have the correct model?". 
On the other hand, we present a hypothetical example to illustrate the Bayesian
workflow presented in the main text.


\tableofcontents 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(bayesplot)
library(colorspace)
library(cmdstanr)
library(dplyr)
library(extraDistr)
library(GGally)
library(ggplot2)
library(ggpubr)
library(kableExtra)
library(MCMCpack)
library(Metrics)
library(parallel)
library(patchwork)
library(posterior)
library(purrr)
library(readsdr)
library(rstan)
library(scales)
library(stringr)
library(tictoc)
library(tidyr)
source("./R/helpers.R")
source("./R/plots.R")
```

\newpage

# Synthetic data

We hypothetically received data (Figure 1) from an outbreak of influenza (H1N1)
in Cumberland, Maryland. This outbreak occurred in 1918, when the population 
size was 5234. From previous studies, we estimate the latent and infectious
periods at 2 days each.

```{r}
mdl_path <- "./models/SEIR.stmx"


N          <- 5234
I_0        <- 2
R_0        <- 1570
stock_list <- list(S = N - I_0 - R_0,
                   E = 0,
                   I = I_0,
                   R = R_0,
                   C = I_0)

const_list <- list(rho = 0.77,
                   par_beta = 1.29,
                   N = N,
                   sigma = 0.5,
                   par_gamma = 0.5,
                   I0 = I_0,
                   R0 = R_0)

actual_pars <- data.frame(name = c("beta", "I(0)", "R(0)", "rho") , 
                          val = c(1.29, I_0, R_0, 0.77))

mdl       <- read_xmile(mdl_path, stock_list = stock_list,
                        const_list = const_list)
```

```{r}
sim_output <- sd_simulate(mdl$deSolve_components, start_time = 0,
                          stop_time = 91, timestep = 1 / 128,
                          integ_method = "rk4")

set.seed(1924)

sim_inc <- sd_net_change(sim_output, "C") %>% dplyr::select(-var) %>% 
  rename(true_inc = value) %>% 
  mutate(measured_inc = rpois(nrow(.), true_inc))

actual_df <- sim_inc %>% 
  rename(x = true_inc,
         y = measured_inc)
```

```{r, fig.cap = "Synthetic incidence", fig.height = 3}
ggplot(sim_inc, aes(x = time, y = true_inc)) +
  geom_line(colour = "grey95") +
  geom_point(data = sim_inc, aes(y = measured_inc), colour = "steelblue") +
  labs(x = "Days since the first case reported",
       y = "Incidence [New cases per day]") +
  theme_pubr()
```

# Calibration

The goal of this exercise is the same as the one described in the main text. In
particular, we aim to estimate the basic reproduction number ($R_0$). To do so, 
we fit, through Hamiltonian Monte Carlo, an SEIR model (see equations in
main text section 2) to the received time-series (incidence data). In doing so,
we can estimate parameter values that best describe the given data, and from 
which we can approximate $R_0$.


```{r}
n_sims    <- 500
pars_list <- vector(mode = "list", length = 3)
mase_list <- vector(mode = "list", length = 3)
```


## Case 1

After formulating a dynamic hypothesis, the next step in a calibration process 
consists of determining which parameters are fixed within the dynamic structure,
and which will be estimated using the MCMC algorithm. We start by considering 
the case where the effective contact rate ($\beta$) and the initial infectious
($I(0)$) are unknown. Further, we assume that there is no underreporting 
($\rho = 1$), and that there are no people with previous immunity ($R(0) = 0$).
We adopt the priors described in the main text.



### Priors


#### Prior distributions

* $\beta \sim lognormal(0,1)$
* $I(0)  \sim lognormal(0,1)$

#### Unmodelled predictors

\hfill

```{r}
data.frame(Predictor = c("$\\rho$", "$\\sigma$", "$\\gamma$", "E(0)", "R(0)", "N"),
           Value     = c(1, 0.5, 0.5, 0, 0, 5234)) -> df

kable(df, booktab = TRUE, escape = FALSE)
```

### Prior predictive checks

Then, we check that our prior information captures the data (Figure 2). Based on
the results below, we *accept the prior*, and proceed to fit the model.

```{r case_1_prior_pred_checks}
fldr      <- "./backup_objs/Synthetic_data/Case_1"
file_path <- file.path(fldr, "prior_sims.rds")

if(!file.exists(file_path)) {
  stock_list <- list(E = 0,
                     R = 0)

  const_list <- list(N         = N,
                     sigma     = 0.5,
                     par_gamma = 0.5,
                     rho       = 1)
  set.seed(2044)
  beta_vals <- rlnorm(n_sims)
  I0_vals   <- rlnorm(n_sims)

  mdl2 <- read_xmile(mdl_path, stock_list = stock_list,
                     const_list = const_list)

  consts_df <- data.frame(par_beta = beta_vals)
  stocks_df <- data.frame(I = I0_vals,
                          S = N - I0_vals,
                          C = I0_vals)

  sens_o <- sd_sensitivity_run(mdl2$deSolve_components, start_time = 0, 
                               stop_time = 91, timestep = 1 / 32,
                               multicore = TRUE, n_cores = 4, 
                               integ_method = "rk4", stocks_df = stocks_df,
                               consts_df = consts_df)
  saveRDS(sens_o, file_path)
} else {
  sens_o <- readRDS(file_path)
}

sens_inc <- predictive_checks(n_sims, sens_o)
```

```{r, fig.cap = "Case 1.Prior predictive checks", fig.height = 3}
meas_df <- sim_inc %>% rename(y = measured_inc) 
plot_predictive_checks(sens_inc, meas_df)
```

### Diagnostics

After validating the prior, we run four chains of 1,000 samples allocated
to each phase (_warm-up_ and _sampling_).


```{r}
pars_hat  <- c("I0", "beta")
gamma     <- 0.5
sigma     <- 0.5
consts    <- sd_constants(mdl)
ODE_fn    <- "SEIR"
stan_fun  <- stan_ode_function(mdl_path, ODE_fn, pars = consts$name[2], 
                               const_list = list(N = N, par_gamma = gamma,
                                                 sigma = sigma, rho = 1))

fun_exe_line <- str_glue("  o = ode_rk45({ODE_fn}, x0, t0, ts, params);") 

stan_data <- stan_data("y", type = "int", inits = FALSE)

stan_params <- paste(
  "parameters {",
  "  real<lower = 0> beta;",
  "  real<lower = 0> I0;",
"}", sep = "\n")

stan_tp <- paste(
  "transformed parameters{",
  "  vector[n_difeq] o[n_obs]; // Output from the ODE solver",
  "  real x[n_obs];",
  "  vector[n_difeq] x0;",
  "  real params[n_params];",
  "  x0[1] = 5234 - I0;",
  "  x0[2] = 0;",
  "  x0[3] = I0;",
  "  x0[4] = 0;",
  "  x0[5] = I0;",
  "  params[1] = beta;",
  fun_exe_line,
  "  x[1] =  o[1, 5]  - x0[5];",
  "  for (i in 1:n_obs-1) {",
  "    x[i + 1] = o[i + 1, 5] - o[i, 5] + 1e-5;",
  "  }",
  "}", sep = "\n")

stan_model <- paste(
  "model {",
  "  beta  ~ lognormal(0, 1);",
  "  I0    ~ lognormal(0, 1);",
  "  y     ~ poisson(x);", 
  "}",
  sep = "\n")

stan_text   <- paste(stan_fun, stan_data, stan_params,
                       stan_tp, stan_model, sep = "\n")

stan_filepath <- "./Stan_files/Synthetic_data/Case_1.stan"

create_stan_file(stan_text, stan_filepath)
```

```{r fit_case_1, results = 'hide'}
fldr       <- "./backup_objs/Synthetic_data/Case_1"
file_path  <- file.path(fldr, "Case_1.rds")

if(!file.exists(file_path)) {
  stan_d <- list(n_obs    = nrow(sim_inc),
                 y        = sim_inc$measured_inc,
               n_params = 1,
               n_difeq  = 5,
               t0       = 0,
               ts       = 1:length(sim_inc$measured_inc))
  
  mod <- cmdstan_model(stan_filepath)

  fit <- mod$sample(data            = stan_d,
                    seed            = 276203775,
                    chains          = 4,
                    parallel_chains = 4,
                    iter_warmup     = 1000,
                    iter_sampling   = 1000,
                    refresh         = 5,
                    save_warmup     = TRUE,
                    output_dir      = fldr)

  fit$save_object(file_path)
} else {
  fit <- readRDS(file_path)
}
```

#### HMC Diagnostics

\hfill

HMC diagnostics are satisfactory and indicate no pathological behaviour.


```{r, message = TRUE}
# We generated this file from $fit$cmdstan_diagnose()
fileName <- file.path(fldr, "diag_1.txt")

readChar(fileName, file.info(fileName)$size) %>% cat()
```


#### Trace plots

\hfill

Furthermore, a visual inspection on trace plots (Figure 3) indicate that the
chains converge.

```{r, fig.height = 3, fig.cap = "Case 1. Trace plots"}
mcmc_trace(fit$draws(), pars_hat)
```

#### Metrics

\hfill

Finally, $\widehat{R}$ and $ESS_{bulk}$ are within the predetermined 
thresholds (< 1.05 and > 400, respectively). Based on all of these diagnostics,
we *accept the computation* and move to the *posterior predictive checking *
stage. 

\hfill

```{r}
par_matrices <- lapply(pars_hat, function(par) {
  extract_variable_matrix(fit$draws(), par)
})

ess_vals   <- map_dbl(par_matrices, ess_bulk)
r_hat_vals <- map_dbl(par_matrices, rhat)

smy_df  <- data.frame(par      = c("I(0)", "$\\beta$"),
                      ess_bulk = ess_vals,
                      rhat     = r_hat_vals)
```

```{r, echo = FALSE}
colnames(smy_df) <- c("par", "$ESS_{bulk}$", "$\\widehat{R}$")
kable(smy_df, booktab = TRUE, escape = FALSE)
```


### Posterior predictive checking

However, when we compare the model's fit against the data (Figure 4), we observe
that the estimated trajectories overestimate the actual behaviour. Assuming no 
underreporting and no previous recovered individuals produces an inadequate
explanation of the observed data. In other words, it does not pass the stringent
validity test of model calibration. For this reason, we deem the model 
*not trustworthy*. 

```{r case1_fit, fig.height = 3, fig.cap = "Case 1. Fit"}
posterior_df <- fit$draws() %>% as_draws_df()

plot_ts_fit(posterior_df, actual_df)
```

```{r case1_MASE}
source("./R/Metrics_utils.R")

mase_list[[1]] <- get_mase_fit(posterior_df, sim_inc$measured_inc) %>% 
  mutate(Case = "1")

pars_df <- posterior_df[, pars_hat]

pars_list[[1]] <- pars_df %>% mutate(Case = "1",
                                     iter = row_number())
```


\newpage


## Case 2

Since Case 1's setup do not yield satisfactory results, we consider
a more complex model. Specifically, $R(0)$ and $\rho$ are no longer assumed as
unmodelled predictors, resulting in a four-unknowns model. We adopt the priors
described in the main text. For $R(0)$, we choose a [weakly informative prior](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations).

### Priors


#### Prior distributions

* $\beta \sim lognormal(0,1)$
* $\rho \sim Beta(2,2)$
* $I(0)  \sim lognormal(0,1)$
* $R(0)  \sim Normal(0,1e4)$

#### Unmodelled predictors

\hfill

```{r}
data.frame(Predictor = c("$\\sigma$", "$\\gamma$", "E(0)", "N"),
           Value     = c(0.5, 0.5, 0, 5234)) -> df

kable(df, booktab = TRUE, escape = FALSE)
```

### Prior predictive checks

Again, we check that our prior information captures the data (Figure 5). Based
on the results below, we *accept the prior*, and proceed to fit the model.

```{r case_2_prior_pred_checks}
fldr      <- "./backup_objs/Synthetic_data/Case_2"
file_path <- file.path(fldr, "prior_sims.rds")

if(!file.exists(file_path)) {
  stock_list <- list(E = 0)

  const_list <- list(N         = N,
                     sigma     = 0.5,
                     par_gamma = 0.5)
  set.seed(3181)
  beta_vals <- rlnorm(n_sims)
  rho_vals  <- rbeta(n_sims, 2, 2)
  R0_vals   <- rhnorm(n_sims, 1e4)
  I0_vals   <- rlnorm(n_sims)
  
  # This validations prevents that there are more infected individuals than 
  # population size
  val_sum   <- R0_vals + I0_vals
  val_idx   <- which(val_sum < 5234)
  
  beta_vals <- beta_vals[val_idx]
  rho_vals  <- rho_vals[val_idx]
  R0_vals   <- R0_vals[val_idx]
  I0_vals   <- I0_vals[val_idx]

  mdl2 <- read_xmile(mdl_path, stock_list = stock_list,
                     const_list = const_list)

  consts_df <- data.frame(par_beta = beta_vals,
                          rho      = rho_vals)
  stocks_df <- data.frame(I = I0_vals,
                          S = N - I0_vals - R0_vals,
                          C = I0_vals,
                          R = R0_vals)

  sens_o <- sd_sensitivity_run(mdl2$deSolve_components, start_time = 0, 
                               stop_time = 91, timestep = 1 / 32,
                               multicore = TRUE, n_cores = 4, 
                               integ_method = "rk4", stocks_df = stocks_df,
                               consts_df = consts_df)
  saveRDS(sens_o, file_path)
} else {
  sens_o <- readRDS(file_path)
}

sens_inc <- predictive_checks(n_sims, sens_o)
```

```{r, fig.cap = "Case 2. Prior predictive checks", fig.height = 3}
meas_df <- sim_inc %>% rename(y = measured_inc) 
plot_predictive_checks(sens_inc, meas_df)
```

```{r}
pars_hat  <- c("I0", "R0", "beta", "rho")
gamma     <- 0.5
sigma     <- 0.5
consts    <- sd_constants(mdl)
ODE_fn    <- "SEIR"
stan_fun  <- stan_ode_function(mdl_path, ODE_fn, pars = consts$name[c(2, 1)], 
                               const_list = list(N = N, par_gamma = gamma,
                                                 sigma = sigma))

fun_exe_line <- str_glue("  o = ode_rk45({ODE_fn}, x0, t0, ts, params);") 

stan_data <- stan_data("y", type = "int", inits = FALSE)

stan_params <- paste(
  "parameters {",
  "  real<lower = 0> beta;",
  "  real<lower = 0, upper = 1> rho;",
  "  real<lower = 0> I0;",
  "  real<lower = 0> R0;",
"}", sep = "\n")

stan_tp <- paste(
  "transformed parameters{",
  "  vector[n_difeq] o[n_obs]; // Output from the ODE solver",
  "  real x[n_obs];",
  "  vector[n_difeq] x0;",
  "  real params[n_params];",
  "  x0[1] = 5234 - I0 - R0;",
  "  x0[2] = 0;",
  "  x0[3] = I0;",
  "  x0[4] = R0;",
  "  x0[5] = I0;",
  "  params[1] = beta;",
  "  params[2] = rho;",
  fun_exe_line,
  "  x[1] =  o[1, 5]  - x0[5];",
  "  for (i in 1:n_obs-1) {",
  "    x[i + 1] = o[i + 1, 5] - o[i, 5] + 1e-4;",
  "  }",
  "}", sep = "\n")

stan_model <- paste(
  "model {",
  "  beta  ~ lognormal(0, 1);",
  "  rho   ~ beta(2, 2);",
  "  I0    ~ lognormal(0, 1);",
  "  R0    ~ normal(0, 1e4);",
  "  y     ~ poisson(x);", 
  "}",
  sep = "\n")

stan_text   <- paste(stan_fun, stan_data, stan_params,
                       stan_tp, stan_model, sep = "\n")

stan_filepath <- "./Stan_files/Synthetic_data/Case_2.stan"

create_stan_file(stan_text, stan_filepath)
```

```{r}
stan_d <- list(n_obs    = nrow(sim_inc),
               y        = sim_inc$measured_inc,
               n_params = 2,
               n_difeq  = 5,
               t0       = 0,
               ts       = 1:length(sim_inc$measured_inc))
```


### First calibration attempt

\hfill

Once we validate the prior, we run four chains of 1,000 samples allocated
to each phase (_warm-up_ and _sampling_).

```{r, results = 'hide'}
fldr       <- "./backup_objs/Synthetic_data/Case_2/Case_2.1"
file_path  <- file.path(fldr, "Case_2.1.rds")

if(!file.exists(file_path)) {

  mod <- cmdstan_model(stan_filepath)

  fit <- mod$sample(data            = stan_d,
                    seed            = 638588660,
                    chains          = 4,
                    parallel_chains = 4,
                    iter_warmup     = 1000,
                    iter_sampling   = 1000,
                    refresh         = 5,
                    save_warmup     = TRUE,
                    output_dir      = fldr)
  
  fit$save_object(file_path)
} else {
  fit <- readRDS(file_path)
}
```


#### HMC Diagnostics

\hfill

Stan informs that there are 47 divergent iterations, an indication of
pathological behaviour that may lead to biased results.

```{r, message = TRUE}
# We generated this file from $fit$cmdstan_diagnose()
fileName <- file.path(fldr, "diag_2.1.txt")

readChar(fileName, file.info(fileName)$size) %>% 
  cat()
```

#### Pairs-plot

\hfill

Thanks to the _bayesplot_ package, we can pinpoint the divergent iterations 
(red crosses) in the parameter space (Figure 6).

```{r, fig.cap = "Case 2.1. Pairs-plot"}
np_cp <- nuts_params(fit)
mcmc_pairs(fit$draws(), np = np_cp, pars = pars_hat,
           off_diag_args = list(size = 0.75))
```


#### Trace plots

\hfill

Interestingly, at first sight (Figure 7),  chains appear to converge. There are
no obvious trends and they seem to mix.

```{r, fig.cap = "Case 2.2. Trace plots", fig.height = 3.5}
mcmc_trace(fit$draws(), pars_hat)
```

Upon further inspection (Figure 8), though, we identify the following issues in
$\beta$'s chains:

- *Chain 2*: Around sample 350, the chain is stuck. The chain around that 
location (~1.4) is not jumping up and down. On the contrary, it stays there for
some iterations.

- *Chain 3*: Around sample 430, the chain is stuck. The chain around that
point (~0.9) is not jumping up and down. On the contrary, it stays there for
some iterations.

- *Chain 4*: Unlike the other three chains, it describe a different 
distribution. If we look at the _center_ of the trace plot, we notice that it is 
different from the other _centers_. Furthermore, the trace plot is narrower than
the other three chains.

```{r, fig.height = 3.5, fig.cap = "Case 2.1. Beta's trace plots"}
beta_draws <- extract_variable_matrix(fit$draws(), "beta") %>% 
  as.data.frame() %>% 
  mutate(id = row_number()) %>% 
  pivot_longer(-id, names_to = "Chain")

ggplot(beta_draws, aes(x = id, y = value)) +
  geom_line(aes(colour = Chain)) +
  scale_colour_manual(values = sequential_hcl(7, palette = "Blues 3")[1:4]) +
  facet_wrap(~Chain) +
  labs(subtitle = "Effective contact rate's chains") +
  theme_pubr()
```

#### Metrics

\hfill

We confirm these findings with $\widehat{R}$ and $ESS_{bulk}$. For two out of
the four parameters, $\widehat{R}$ is higher than its recommended
threshold (1.05), indicating that convergence was not achieved. Likewise, no
$ESS_{bulk}$ is above the 400 threshold. Consequently, we deem the computation 
as *not valid*.

```{r}
par_matrices <- lapply(pars_hat, function(par) {
  extract_variable_matrix(fit$draws(), par)
})

ess_vals   <- map_dbl(par_matrices, ess_bulk)
r_hat_vals <- map_dbl(par_matrices, rhat)

smy_df  <- data.frame(par      = c("I(0)", "R(0)", "$\\beta$", "$\\rho$"),
                      ess_bulk = ess_vals,
                      rhat     = r_hat_vals)
```

```{r, echo = FALSE}
colnames(smy_df) <- c("par", "$ESS_{bulk}$", "$\\widehat{R}$")
kable(smy_df, booktab = TRUE, escape = FALSE)
```


### Second calibration attempt

According to the [Stan manual](https://mc-stan.org/misc/warnings.html), one 
strategy to address divergent transitions is increasing the _adapt_delta_ 
parameter. This is the target average proposal acceptance probability during
Stan’s adaptation period, and increasing it will force Stan to take smaller
steps. We increase this parameter from its default value (0.8) to 0.99, and
re-run the sampling process.

```{r, results = 'hide'}
fldr       <- "./backup_objs/Synthetic_data/Case_2/Case_2.2"
file_path  <- file.path(fldr, "Case_2.2.rds")

if(!file.exists(file_path)) {

  mod <- cmdstan_model(stan_filepath)

  fit <- mod$sample(data            = stan_d,
                    seed            = 638588660,
                    chains          = 4,
                    parallel_chains = 4,
                    iter_warmup     = 1000,
                    iter_sampling   = 1000,
                    adapt_delta     = 0.99,
                    refresh         = 5,
                    save_warmup     = TRUE,
                    output_dir      = fldr)
  
  fit$save_object(file_path)
} else {
  fit <- readRDS(file_path)
}
```

#### HMC Diagnostics

\hfill

By increasing _adapt_delta_, we eliminated the divergent transitions at the cost
of hitting the maximum treedepth threshold, which unlike divergent transitions
is not a validity concern but an efficiency indicator. Despite this indicator
inform us about the complexity of the posterior explored, we could still use
the samples for further analysis if the converge and efficiency metrics are
satisfactory.


```{r, message = TRUE}
# We generated this file from $fit$cmdstan_diagnose()
fileName <- file.path(fldr, "diag_2.2.txt")

readChar(fileName, file.info(fileName)$size) %>% 
  cat()
```

#### Pairs-plot

\hfill

In this updated pairs-plot (Figure 9), we highlight the iterations where the 
algorithm hit the maximum treedepth, that is, the maximum allowed number of 
steps taken within an iteration. 

```{r, fig.cap = "Case 2.2. Pairs-plot"}
np_cp <- nuts_params(fit)
mcmc_pairs(fit$draws(), np = np_cp, pars = pars_hat,
           off_diag_args = list(size = 0.75),
           max_treedepth = 9)
```

#### Trace plots

\hfill

Trace plots (Figure 10) do not reveal obvious issues in convergence and efficiency.

```{r, fig.cap = "Case 2.2. Trace plots", fig.height = 3.5}
mcmc_trace(fit$draws(), pars_hat)
```

#### Metrics

\hfill

$\widehat{R}$ confirms the visual inspection from the trace plots. The chains do
converge. However, hitting the maximum treedepth affected efficiency. 
$ESS_{bulk}$ (below 400) indicates that there is no sufficient information to 
estimate quantities of interest.

```{r}
par_matrices <- lapply(pars_hat, function(par) {
  extract_variable_matrix(fit$draws(), par)
})

ess_vals   <- map_dbl(par_matrices, ess_bulk)
r_hat_vals <- map_dbl(par_matrices, rhat)

smy_df  <- data.frame(par      = c("I(0)", "R(0)", "$\\beta$", "$\\rho$"),
                      ess_bulk = ess_vals,
                      rhat     = r_hat_vals)
```

```{r, echo = FALSE}
colnames(smy_df) <- c("par", "$ESS_{bulk}$", "$\\widehat{R}$")
kable(smy_df, booktab = TRUE, escape = FALSE)
```

### Third calibration attempt

To address the efficiency issues detected in the previous run, we increase
in the number of samples from 1,000 to 1,500, and increase the maximum number
of steps allowed per iteration.

```{r, results = 'hide'}
fldr       <- "./backup_objs/Synthetic_data/Case_2/Case_2.3"
file_path  <- file.path(fldr, "Case_2.3.rds")

if(!file.exists(file_path)) {

  mod <- cmdstan_model(stan_filepath)

  fit <- mod$sample(data            = stan_d,
                    seed            = 638588660,
                    chains          = 4,
                    parallel_chains = 4,
                    iter_warmup     = 1500,
                    iter_sampling   = 1500,
                    adapt_delta     = 0.99,
                    max_treedepth   = 20,
                    refresh         = 5,
                    save_warmup     = TRUE,
                    output_dir      = fldr)
  
  fit$save_object(file_path)
} else {
  fit <- readRDS(file_path)
}
```

#### HMC Diagnostics

\hfill

HMC diagnostics are satisfactory and indicate no pathological behaviour.

```{r, message = TRUE}
# We generated this file from $fit$cmdstan_diagnose()
fileName <- file.path(fldr, "diag_2.3.txt")

readChar(fileName, file.info(fileName)$size) %>% 
  cat()
```

#### Trace plots

\hfill

Trace plots (Figure 11) do not reveal obvious issues in convergence and 
efficiency.

```{r, fig.height = 3, fig.cap = "Case 2.3. Trace plots"}
mcmc_trace(fit$draws(), pars_hat)
```

#### Metrics

\hfill

$\widehat{R}$ and $ESS_{bulk}$ are within the predetermined thresholds
(< 1.05 and > 400, respectively). Based on all of these diagnostics,
we *accept the computation* and move forward to the
*posterior predictive checking * stage. 

\hfill

```{r}
par_matrices <- lapply(pars_hat, function(par) {
  extract_variable_matrix(fit$draws(), par)
})

ess_vals   <- map_dbl(par_matrices, ess_bulk)
r_hat_vals <- map_dbl(par_matrices, rhat)



smy_df  <- data.frame(par      = c("I(0)", "R(0)", "$\\beta$", "$\\rho$"),
                      ess_bulk = ess_vals,
                      rhat     = r_hat_vals)
```

```{r, echo = FALSE}
colnames(smy_df) <- c("par", "$ESS_{bulk}$", "$\\widehat{R}$")
kable(smy_df, booktab = TRUE, escape = FALSE)
```

### Posterior predictive checking

By comparing simulated trajectories and the actual data (Figure 12), we see that
this structure is an adequate explanation of the observed incidence. Thus, we 
_accept the model_ for parameter estimation.

```{r case2_fit, fig.height = 3, fig.cap = "Case 2. Fit"}
posterior_df <- fit$draws() %>% as_draws_df()

plot_ts_fit(posterior_df, actual_df)
```

### Parameter estimations

Before estimating marginal posterior uncertainty intervals, it is good practice
to check for the correlations among parameters (Figure 13). Here, we can 
appreciate this model's complexity by looking at the large degree of 
correlations in this parameter space.

```{r case_2_distributions, fig.cap = "Case 2. Marginal and joint distributions"}
pars_df <- posterior_df[, pars_hat]
pairs_posterior(pars_df, strip_text = 10)

pars_list[[2]] <- pars_df %>% mutate(Case = "2",
                                     iter = row_number())
```

### Uncertainty intervals

```{r}
pars_df %>% mutate(id = row_number()) %>% 
  pivot_longer(-id) %>% 
  group_by(name) %>% 
  summarise(Mean       = round(mean(value), 2),
            bound_val  = round(quantile(value, c(0.025, 0.975)),2),
            bound_type = c("2.5\\%", "97.5\\%")) %>% 
  pivot_wider(names_from = bound_type, values_from = bound_val) -> est_df
```

```{r, echo = FALSE}
est_df$name <- c("$\\beta$", "I(0)", "R(0)","$\\rho$")
kable(est_df, booktab = TRUE, escape = FALSE)
```

```{r case2_MASE}
source("./R/Metrics_utils.R")

mase_list[[2]] <- get_mase_fit(posterior_df, sim_inc$measured_inc) %>% 
  mutate(Case = "2")
```



## Case 3

Although Case 2 is computationally satisfactory, the uncertainty around the
estimates could be improved if there were more data. Fortunately, a new 
pre-print has just been published. This study reports that 30 % of the 
population have antibodies against this virus. In light of this new data, we
update the calibration setup.

### Priors

#### Prior distributions

* $\beta \sim lognormal(0,1)$
* $\rho \sim Beta(2,2)$
* $I(0)  \sim lognormal(0,1)$

#### Unmodelled predictors

\hfill

```{r}
data.frame(Predictor = c("$\\sigma$", "$\\gamma$", "E(0)", "R(0)", "N"),
           Value     = c(0.5, 0.5, 0, 1570, 5234)) -> df

kable(df, booktab = TRUE, escape = FALSE)
```

### Prior predictive checks

As in Cases 1 & 2, we check that our prior information captures the data. Based
on the results (Figure 14), we *accept the prior*, and proceed to fit the model.

```{r case_3_prior_pred_checks}
fldr      <- "./backup_objs/Synthetic_data/Case_3"
file_path <- file.path(fldr, "prior_sims.rds")

if(!file.exists(file_path)) {
  
  stock_list <- list(E = 0,
                     R = 1570)

  const_list <- list(N         = N,
                     sigma     = 0.5,
                     par_gamma = 0.5)
  
  set.seed(2041)
  beta_vals <- rlnorm(n_sims)
  rho_vals  <- rbeta(n_sims, 2, 2)
  I0_vals   <- rlnorm(n_sims)
  
  mdl2 <- read_xmile(mdl_path, stock_list = stock_list,
                     const_list = const_list)

  consts_df <- data.frame(par_beta = beta_vals,
                          rho      = rho_vals)
  
  stocks_df <- data.frame(I = I0_vals,
                          S = N - I0_vals - 1570,
                          C = I0_vals)

  sens_o <- sd_sensitivity_run(mdl2$deSolve_components, start_time = 0, 
                               stop_time = 91, timestep = 1 / 32,
                               multicore = TRUE, n_cores = 4, 
                               integ_method = "rk4", stocks_df = stocks_df,
                               consts_df = consts_df)
  saveRDS(sens_o, file_path)
} else {
  sens_o <- readRDS(file_path)
}

sens_inc <- predictive_checks(n_sims, sens_o)
```

```{r, fig.cap = "Case 3. Prior predictive checks", fig.height = 3}
meas_df <- sim_inc %>% rename(y = measured_inc) 
plot_predictive_checks(sens_inc, meas_df)
```

### Diagnostics

In this calibration setup, we run four chains of 1,000 samples allocated
to each phase (_warm-up_ and _sampling_) and HMC's default settings. 


```{r}
pars_hat  <- c("I0", "beta", "rho")

stan_filepath <- "./Stan_files/example/flu_poisson.stan"
```

```{r case_3_run, results = 'hide'}
fldr       <- "./backup_objs/Synthetic_data/Case_3"
file_path  <- file.path(fldr, "Case_3.rds")

if(!file.exists(file_path)) {
  stan_d <- list(n_obs    = nrow(sim_inc),
                 y        = sim_inc$measured_inc,
               n_params = 2,
               n_difeq  = 5,
               t0       = 0,
               ts       = 1:length(sim_inc$measured_inc))

mod <- cmdstan_model(stan_filepath)

fit <- mod$sample(data            = stan_d,
                  seed            = 594675061,
                  chains          = 4,
                  parallel_chains = 4,
                  iter_warmup     = 1000,
                  iter_sampling   = 1000,
                  refresh         = 5,
                  save_warmup     = TRUE,
                  output_dir      = fldr)

  fit$save_object(file_path)
} else {
  fit <- readRDS(file_path)
}
```

#### HMC Diagnostics

\hfill

HMC diagnostics are satisfactory and indicate no pathological behaviour.


```{r case_3_HMC_diag, message = TRUE}
# We generated this file from $fit$cmdstan_diagnose()
fileName <- file.path(fldr, "diag_3.txt")

readChar(fileName, file.info(fileName)$size) %>% cat()
```

### Trace plot

\hfill

Trace plots (Figure 15) do not reveal obvious issues in convergence and 
efficiency.

```{r case_3_trace_plots, fig.cap = "Case 3. Trace plots", fig.height = 2.5}
mcmc_trace(fit$draws(), pars_hat)
```

#### Metrics

\hfill

$\widehat{R}$ and $ESS_{bulk}$ are within the predetermined thresholds
(< 1.05 and > 400, respectively). Based on all of these diagnostics,
we *accept the computation* and move to the *posterior predictive checking *
stage. 

\hfill

```{r}
par_matrices <- lapply(pars_hat, function(par) {
  extract_variable_matrix(fit$draws(), par)
})

ess_vals   <- map_dbl(par_matrices, ess_bulk)
r_hat_vals <- map_dbl(par_matrices, rhat)



smy_df  <- data.frame(par      = c("I(0)", "$\\beta$", "$\\rho$"),
                      ess_bulk = ess_vals,
                      rhat     = r_hat_vals)
```

```{r, echo = FALSE}
colnames(smy_df) <- c("par", "$ESS_{bulk}$", "$\\widehat{R}$")
kable(smy_df, booktab = TRUE, escape = FALSE)
```

### Posterior predictive checking

By comparing simulated trajectories and the actual data (Figure 16), we see that
this  structure is an adequate explanation of the observed incidence. Thus, we 
_accept the model_ for parameter estimation.

```{r case3_fit, fig.height = 3, fig.cap = "Case 3. Fit"}
posterior_df <- fit$draws() %>% as_draws_df()

plot_ts_fit(posterior_df, actual_df)
```

### Parameter estimations

As with Case 2, we check for the correlations among parameters (Figure 17). Even
though the correlation between $I(0)$ and $\beta$ slightly increased; 
correlations between $\rho$ and $I(0)$, and $\rho$ and $\beta$ were
significantly reduced in this revised setup.


```{r case_3_distributions, fig.cap = "Case 3. Marginal and joint distributions"}
pars_df <- posterior_df[, pars_hat]
pairs_posterior(pars_df, strip_text = 10)

pars_list[[3]] <- pars_df %>% mutate(Case = "3",
                                     iter = row_number())
```


### Uncertainty intervals

Finally, we accept these estimates as we deem they provide useful information.

```{r}
pars_df %>% mutate(id = row_number()) %>% 
  pivot_longer(-id) %>% 
  group_by(name) %>% 
  summarise(Mean       = round(mean(value), 2),
            bound_val  = round(quantile(value, c(0.025, 0.975)),2),
            bound_type = c("2.5\\%", "97.5\\%")) %>% 
  pivot_wider(names_from = bound_type, values_from = bound_val) -> est_df
```

```{r, echo = FALSE}
est_df$name <- c("$\\beta$", "I(0)", "$\\rho$")
kable(est_df, booktab = TRUE, escape = FALSE)
```

```{r case3_MASE}
source("./R/Metrics_utils.R")

mase_list[[3]] <- get_mase_fit(posterior_df, sim_inc$measured_inc) %>% 
  mutate(Case = "3")
```

## Fit performance

To conclude with this section, we present a comparison of the calibration 
performance among the three cases.

### MASE 


Similar to the example presented in the main text, we use the Mean Absolute 
Scaled Error (MASE) as a measure of fit accuracy. The reader should recall that
MASE < 1 indicates adequate performance, and lower values point to better fits.
In Figure 18A, it is seen that unlike Cases 2 and 3, Case 1 does not explain the 
data adequately. Moreover, Cases 2 and 3 obtained similar performance scores
(Figure 18B).

```{r, fig.height = 4, fig.width = 4, fig.cap = "MASE comparison"}
all_MASE_df <- do.call("bind_rows", mase_list)

ggplot(all_MASE_df, aes(x = Case, y = mase)) +
  geom_boxplot(aes(group = Case)) +
  geom_hline(yintercept = 1, colour = "red", linetype ="dotted") +
  theme_pubr() +
  labs(subtitle = "A)", y = "MASE")-> g1

ggplot(all_MASE_df %>% 
         filter(Case != "1"), aes(x = Case, y = mase)) +
  geom_boxplot(aes(group = Case)) +
  geom_hline(yintercept = 1, colour = "red", linetype ="dotted") +
  theme_pubr() +
  labs(subtitle = "B)", y = "MASE")-> g2

g1 / g2
```




### Parameter recovery

Lastly, in Figure 19, we present posterior marginal distributions compared 
against the actual value (red dashed line). In Case 1, failing to explain the 
data adequately leads to obtain unreliable estimates. Conversely, both Cases 2 &
3 explains correctly the data and capture the actual values. However, they 
differ in the degree of uncertainty.


```{r, fig.cap = "Parameter estimates per case", fig.height = 3.5}
all_df <- do.call("bind_rows", pars_list)

tidy_all <- all_df %>% pivot_longer(c(-Case, -iter)) %>% 
  mutate(name = case_when(
    name == "I0" ~ "I(0)",
    name == "R0" ~ "R(0)",
    TRUE ~ as.character(name)))

ggplot(tidy_all, aes(x = Case, y = value)) +
  geom_boxplot(aes(group = Case)) +
  facet_wrap(~name, scales = "free", labeller = label_parsed) +
  geom_hline(data = actual_pars, aes(yintercept = val), alpha = 0.5, 
             colour = "red", linetype = "dashed") +
  theme_pubr()
```

\newpage

# Random-Walk Metropolis fit

We add this section to further corroborate that an SEIR with four unknown
parameters is a  high-dimensional structure that generates a complex parameter
space of difficult exploration. To demonstrate this complexity, we fit the
aforementioned model to the synthetic incidence data using the Random-Walk
Metropolis algorithm.


```{r}
source("./R/sd_loglik_fun.R")
source("./R/posterior_components.R")

set.seed(131103)
inits_I    <- runif(4, -2, 2)
inits_R    <- log(runif(4, 0, 5000))
inits_beta <- runif(4, -2, 2)
inits_rho  <- runif(4, -2, 2)

pars_df <- data.frame(name      = c("I", "R", "par_beta", "rho"),
                      type      = c("stock", "stock", "constant", "constant"),
                      par_trans = c("log", "log", "log", "logit"))

sim_controls <- list(start = 0, stop = 91, step = 1 / 64, integ_method = "rk4")

fit_options <- list(stock_name = "C", 
                    stock_fit_type = "net_change",
                    dist = list(name     = "dpois",
                                sim_data = "lambda",
                                dist_offset = "1e-5"),
                    data = sim_inc$measured_inc)

extra_stocks <- list(list(name = "C", init = "I"),
                     list(name = "S", init = "5234 - R - I"))

extra_constraints <- list("I + R >= 5234")

loglik <- sd_loglik_fun(pars_df, mdl$deSolve_components, sim_controls,
                        fit_options, extra_stocks, extra_constraints)

logprior <- function(pars) {
  par_I      <- exp(pars[1])
  par_R      <- exp(pars[2])
  par_beta   <- exp(pars[3])
  par_rho    <- expit(pars[4])
  
  dlnorm(par_I, meanlog = 0, sd = 1, log = TRUE)      +
    dnorm(par_R, mean = 0, sd = 1e4, log = TRUE) +
    dlnorm(par_beta, meanlog = 0, sd = 1, log = TRUE) +
    dbeta(par_rho, shape1 = 2, shape2 = 2, log = TRUE)
}

posterior_fun <- function(pars) loglik(pars) + logprior(pars) 
```

```{r}
fn <- "./backup_objs/Synthetic_data/Case_2_RWM.rds"

if(!file.exists(fn)) {
  tic.clearlog()
  tic()
  
  mclapply(1:4, function(i) {
    n_iter  <- 10000
    inits   <- c(inits_I[[i]], inits_R[[i]], inits_beta[[i]], inits_rho[[i]])
    seed    <- 149694674
    
    MCMCmetrop1R(
      fun = posterior_fun,
      theta.init = inits,
      mcmc       = n_iter, 
      burnin     = n_iter,
      seed       = seed, 
      verbose    = 1) -> mcmc_output
    
  posterior_df <- as.data.frame(mcmc_output) %>%
    set_names(c("I0", "R0", "beta", "rho")) %>% 
    mutate(I0   = exp(I0),
           R0   = exp(R0),
           beta  = exp(beta),
           rho   = expit(rho),
           Chain = i)
  }, mc.cores = 4) -> post_obj

  toc(quiet = FALSE, log = TRUE)
  log.lst <- tic.log(format = FALSE)
      
  result_obj <- list(post_obj     = post_obj,
                     time         = log.lst)
  
  saveRDS(result_obj, fn)
} else {
  result_obj <- readRDS(fn)
}
```

```{r}
caption <- "Unknown parameters' trace plots constructed from the samples obtained from RWM." 
```


## Trace plots

In Figure 20, it can be seen that after 10,000 burn-in iterations (the reader
should recall that the trace plots presented below correspond to the samples
generated in the *sampling phase*), the chains do not converge. On the one
hand, one of the chains (Chain 1) is not stationary considering that it exhibits
visible trends in parameters $I(0)$ (downwards) and $\beta$ (upwards). On the 
other hand, all four chains describe different distributions. Namely, they do
not mix.

```{r, fig.cap = caption}
posterior_df <- do.call("bind_rows", result_obj$post_obj)

color_scheme_set("viridis")
bayesplot::mcmc_trace(posterior_df)
```

## Actual values coverage

Considering that the sampling process does not converge, parameter estimates 
from the samples cannot be trusted. However, for illustration purposes, we 
compare such estimates with actual values (Figure 21). In this comparison, we 
observe the unreliability of the estimated posterior distribution as not all 
chains captured the desired values.

```{r}
caption <- "Marginal posterior distributions per chain obtained from fitting (via RWM) an SEIR model with four unknowns to incidence data"
```


```{r, fig.cap = caption, fig.height = 3.5}
tidy_posterior <- posterior_df %>% 
  rename(`I(0)` = I0,
         `R(0)` = R0) %>% 
  mutate(iter = row_number()) %>% 
  pivot_longer(c(-Chain, - iter))

ggplot(tidy_posterior, aes(x = Chain, y = value)) +
  geom_boxplot(aes(group = Chain)) +
  facet_wrap(~name, scales = "free", labeller = label_parsed) +
  geom_hline(data = actual_pars, aes(yintercept = val), alpha = 0.5, 
             colour = "red", linetype = "dashed") +
  labs(y = "Value") +
  theme_pubr()
```


